Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-06-04 PAPER.tex
minor typo and syntax changes

\begin{eqnarray}
\label{eq:sigloss}
\widehat{P}_{\rm out,\alpha} &\equiv& \widehat{\textbf{P}}_{r,\alpha} \nonumber\\%-\widehat{\textbf{P}}_{x,\alpha} \nonumber \\
&=& \text{M}^{\alpha}_{r}\textbf{r}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{r},% - \text{M}^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}^{\alpha}\textbf{R}_{x}\textbf{x},
\end{eqnarray}
where, for illustrative purposes and notational simplicity, we have written these equations with scalar normalizations M, even though for our numerical results we choose a non-diagonal matrix normalization using $\mathbf{M}$ as in Equation \eqref{eq:phat}.

@@

{\bf In between:}
When the injected signal is of a similar amplitude to the data by itself, the situation becomes less straightforward. We see that 
the weighted injected power spectrum component mirrors the input power indicating little loss (i.e. the green curve follows the dotted black line), eventually 
departing from unity when the injected amplitude is well above the level of the data power spectrum. However, 
in this regime the cross-term has nearly the same amplitude, but with a negative sign. As explained below, this negativity is the result of cross-correlating inverse covariance weighted terms.  This negative component drives down the $\widehat{P}_{\rm out}$ estimator (black). We note that in \citetalias{ali_et_al2015}, signal loss was computed by only looking at the second term in Equation \eqref{eq:crossterm} (green), which incorrectly implies no loss at the data-only power spectrum level. Ignoring the effect of the negative power from the cross-terms is the main reason for under-estimating power spectrum limits in \citetalias{ali_et_al2015}.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/sigloss_modeloop_2panel.pdf}
	\caption{Power spectra $2\sigma$ upper limits for $k=0.393$\,$h$ Mpc$^{-1}$ for fringe-rate filtered PAPER-64 data. Top: Values 
are shown before (dashed red) and after (solid red) signal loss estimation via our signal injection framework as a function of number of eigenmodes of $\widehat{\textbf{C}}$ that 
are down-weighted. This regularization knob is tuned from $0$ modes on the left (i.e. unweighted) to $21$ modes on the right (i.e. the full inverse 
covariance estimator). Over $\sim3$ orders of magnitude of signal loss results when using empirically estimated inverse covariance weighting. Bottom: Power spectrum upper limits before (dashed red) and after (solid red) signal loss estimation as a function of identity added to the empirical covariance. This regularization knob is tuned from $\gamma = 10^{-4}$ on the right (i.e. very little regularization) to $\gamma = 1$ on the left (see main text for the definition of $\gamma$). Also 
plotted in both panels for comparison are $2\sigma$ power spectrum upper limits for the uniform-weighted case (blue) and inverse variance 
weighted case (black); both are after signal loss estimation. Finally, a theoretical prediction for noise ($2\sigma$ error) is plotted 
as green. In the PAPER-64 analysis in this paper, we choose to use a regularization scheme of $\widehat{\textbf{C}}_{\rm eff} \equiv 0.9 \, $Tr($\widehat{\textbf{C}})\textbf{I} + \widehat{\textbf{C}}$ ($\gamma = 0.9$) as a simple example of regularization that minimizes loss, and note that the power spectrum limits using this type of regularization are roughly constant across a large range of values of $\gamma$.}
	\label{fig:sigloss_modeloop}
\end{figure*}

In addition to our thermal noise prediction (green) and uniform-weighted power spectrum limit (blue), one additional horizontal line is shown in Figure \ref{fig:sigloss_modeloop} 
in both panels and represents a third regularization technique. This line (black) denotes the power spectrum value, post-signal loss estimation, for inverse variance weighting (multiplying an identity 
matrix element-wise to $\widehat{\textbf{C}}$). This result is single-valued and not a function of the horizontal axis. We see that all three regularization schemes shown (solid red, dashed red, black) perform similarly at 
their best (i.e. when $\sim2$-$3$ eigenmodes are down-weighted in the case of the solid red curve). However, for the remainder of this paper, we choose to use the weighting option of $\widehat{\textbf{C}} \equiv \widehat{\textbf{C}} + 0.9 \,$Tr($\widehat{\textbf{C}})\textbf{I}$, or $\gamma = 0.9$, which we will denote as $\widehat{\textbf{C}}_{\rm eff}$. We choose this weighting scheme merely as a simple example of regularizing PAPER-64 covariances, noting that the power spectrum upper limit remains roughly constant for a broad range of values of $\gamma$. We also note that all of the regularizations described here perform similarly to the uniformly weighted case. However, we find that the regularization of empirical covariances does outperform the uniform weighted case for redshifts (and $k$-values) for which there are stronger foregrounds, as presented in Kolopanis et al. (\textit{submitted}).

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/ps1_noise_add.pdf}
	\includegraphics[width=0.4\textwidth]{plots/ps2_noise_add.pdf}
	\caption{The power spectrum for a noise simulation that mimics the noise level of a subset of PAPER-64 data, where the solid red curve is the $2\sigma$ upper limit on the EoR signal estimated from our signal injection framework using $\widehat{\textbf{C}}_{\rm eff}$. The theoretical $2\sigma$ thermal noise level prediction based on observational parameters (calculated by Equation \eqref{eq:sense}) is in green. Additionally, the power spectrum results for the uniform weighted case is shown in three different ways: power spectrum values (black and gray points as positive and negative values, respectively, with $2\sigma$ error bars), the $2\sigma$ upper limit on the EoR signal using our full signal injection framework (solid blue), and the measured power spectrum values with $2\sigma$ thermal noise errors (gray shaded regions). The vertical dashed black lines signify the horizon limit for this analysis using $30$\,m baselines. We highlight that the bootstrapped data points and thermal noise prediction show good agreement, while the limits from the full injection framework are inflated due to the additional inclusion of sample variance that comes from the injection simulations.}
	\label{fig:ps_noise}
\end{figure*}

@@

Given our new understanding of the sensitivity of bootstraps to the number of elements sampled, we have removed the second 
bootstrapping step along time entirely and now simply bootstrap over the baseline axis. Power spectrum $2\sigma$ errors (computed from bootstrap variances) with this bootstrapping change for fringe-rate filtered noise are shown in Figure 
\ref{fig:data_errors}. The estimates are uniformly weighted in order to disentangle the effects of bootstrapping from signal loss. As 
shown in the figure, when more elements are drawn for each bootstrap than the number of 
independent samples (by over-sampling elements along the time axis), repeated values begin to crop up and the apparent variation between bootstraps drops, resulting in limits (gray) below the predicted noise level (green). Using the revised bootstrapping method, where bootstrapping only occurs over the baseline axis, the limits (black) are shown to better agree with the analytic prediction for noise. While Figure \ref{fig:data_errors} implies that errors are under-estimated by a factor of $\sim$ $5$ in mK$^{2}$ for the noise simulation, in practice this factor is lower for the case of real data (a factor of $\sim$ $2$ in mK$^{2}$ instead), possibly due to the data being less correlated in time than the fringe-rate filtered noise in the simulation. }%Figure \ref{fig:data_errors} implies that power spectrum upper limits were originally under-estimated by a factor of $\sim$ $5$ in mK$^{2}$.

\item{A second change to our bootstrapping procedure is that we now bootstrap over baseline cross-products, instead of baselines themselves. In the previous analysis, baselines were bootstrapped prior to forming cross power spectra, and using this particular ordering of operations (bootstrapping, then cross-multiplication) yields variances that have been found to disagree with predicted errors from bootstrapping using simulations. On the contrary, bootstrapping over cross power spectra ensures that we are estimating the variance of our quantity of interest (i.e. the power spectrum). This change, while fundamental in retaining the integrity of the bootstrapping method in general, alters the resulting power spectrum errors by factors of $<2$ in practice.}

@@

The second null test possibility (error bars consistent with zero but not with thermal noise) is displayed by the $k$-values just outside the horizon ($k\sim\pm 0.1$\,$h$ Mpc$^{-1}$) for all three tests. This indicates an additive noise component that is increasing our errors. More specifically, although we expect each cross-multiplication that is used in power spectrum estimation to have independent noise, there is still the possibility of a noise-foreground coupling term that can introduce power. This is because cross-multiplications produce four additive terms --- a signal-squared term (where signal includes both foregrounds and EoR), two cross-terms between the signal and noise, and one noise-only term. When differencing two power spectra (each with their own four terms), we expect the signal-only term to subtract out for a ``passing" null test, and we expect the noise-only terms to be consistent with thermal noise. While the cross-terms have a mathematical expectation value of zero, in practice we are limited by our number of samples ($90$ cross-products for this analysis times $\sim$ $8$ independent LST samples). Combined with the fact that the foregrounds are so bright, the finite ensemble of the couplings can introduce extra variance that varies with foreground strength. It is therefore not surprising that this effect is largest at $k$-values just outside the horizon, where we expect foregrounds to be brightest post-delay filtering.

@@

\begin{itemize}
\item Signal loss, previously found to be $<2\%$ in \citetalias{ali_et_al2015}, was under-estimated by a factor of $>$$1000$ for empirically estimated inverse 
covariance weighting. Using a regularized covariance weighting method can minimize loss (Section 
\ref{sec:Weight}), however, because a regularized weighting method is not as aggressive as the former, our results are still a 
factor of $\sim10$ in mK higher than the lossy empirical inverse covariance limits. Under-estimated signal loss therefore represents the bulk of our revision. 
This revision is similar to the re-analysis of results from the GMRT (\citealt{paciga_et_al2013}) which were also 
revised from new signal loss calculations associated with their singular value decomposition foreground filter.
\item Power spectrum errors, originally computed by bootstrapping, were under-estimated for the data by a factor of $\sim2$ in mK due to oversampling data whose effective number of independent samples was reduced from fringe-rate filtering (Section \ref{sec:Boot}). 
\item Several factors used in an analytic expression to predict the noise-level in PAPER-64 data were revised, yielding a 
decrease in predicted sensitivity level by a factor of $\sim3$ in mK (Section \ref{sec:PSSense}). We note that our sensitivity prediction is revised by a factor less than our overall
power spectrum result, implying that if taken at face value, the theoretical prediction for noise in \citetalias{ali_et_al2015} was too high for its data 
points.
\end{itemize}

@@

When we simulate our full injection framework as in Section \ref{sec:Practice}, we sample 50 $P_{\rm in}$ values that range from $\sim$$10^{5}$\,mK$^{2}$ ($h^{-1}$ Mpc)$^{3}$ to $\sim$10$^{11}$\,mK$^{2}$ ($h^{-1}$ Mpc)$^{3}$, and we note that the prior is set to zero outside those regions. For the injections that we do sample, we can easily fit analytic functions to the mean and standard deviations of $\widehat{P}_{\rm out}$ ($\bar y$ and $\sigma$) as functions of $P_{\rm in}$. An example of the typical shape of these functions for the PAPER-64 analysis is shown in Figure \ref{fig:jeffreys1}, though we note in practice we fit solutions for every $k$-value and simulation independently.

@@

\begin{figure}
	\centering
	\includegraphics[width=10cm]{plots/jeffrey_prior.pdf}
	\caption{An example of the typical Jeffreys prior shape for the PAPER-64 analysis as computed by Equation \eqref{eq:jeffreys_final} (black). We smooth the prior using a sliding boxcar average for every $5$ injections (red). Most noticeably, the Jeffreys prior is not constant with $P_{\rm in}$, meaning a uniform prior would be an informative prior.}
	\label{fig:jeffreys2}
\end{figure}

>>>>

\begin{eqnarray}
\label{eq:sigloss}
\widehat{P}_{\rm out,\alpha} &\equiv& \widehat{\textbf{P}}_{r,\alpha} \nonumber\\%-\widehat{\textbf{P}}_{x,\alpha} \nonumber \\
&=& \text{M}^{\alpha}_{r}\textbf{r}^{\dagger}\textbf{R}_{r}\textbf{Q}^{\alpha}\textbf{R}_{r}\textbf{r},% - \text{M}^{\alpha}_{x}\textbf{x}^{\dagger}\textbf{R}_{x}\textbf{Q}^{\alpha}\textbf{R}_{x}\textbf{x},
\end{eqnarray}
where, for illustrative purposes and notational simplicity, we have written these equations with scalar normalizations M, even though for our numerical results we choose a diagonal matrix normalization using $\mathbf{M}$ as in Equation \eqref{eq:phat}.

@@

{\bf In between:}
When the injected signal is of a similar amplitude to the data by itself, the situation becomes less straightforward. We see that 
the weighted injected power spectrum component mirrors the input power indicating little loss (i.e. the green curve follows the dotted black line), eventually 
departing from unity when the injected amplitude is well above the level of the data power spectrum. However, 
in this regime the cross-term (red) has nearly the same amplitude, but with a negative sign. As explained below, this negativity is the result of cross-correlating inverse covariance weighted terms.  This negative component drives down the $\widehat{P}_{\rm out}$ estimator (black). We note that in \citetalias{ali_et_al2015}, signal loss was computed by only looking at the second term in Equation \eqref{eq:crossterm} (green), which incorrectly implies no loss at the data-only power spectrum level. Ignoring the effect of the negative power from the cross-terms is the main reason for under-estimating power spectrum limits in \citetalias{ali_et_al2015}.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1\textwidth]{plots/sigloss_modeloop_2panel.pdf}
	\caption{Power spectra $2\sigma$ upper limits for $k=0.393$\,$h$ Mpc$^{-1}$ for fringe-rate filtered PAPER-64 data. Top: Values 
are shown before (dashed red) and after (solid red) signal loss estimation via our signal injection framework as a function of number of eigenmodes of $\widehat{\textbf{C}}$ that 
are down-weighted. This regularization knob is tuned from $0$ modes on the left (i.e. unweighted) to $21$ modes on the right (i.e. the full inverse 
covariance estimator). Over $\sim3$ orders of magnitude of signal loss results when using empirically estimated inverse covariance weighting. Bottom: Power spectrum upper limits before (dashed red) and after (solid red) signal loss estimation as a function of identity added to the empirical covariance. This regularization knob is tuned from $\gamma = 10^{-4}$ on the right (i.e. very little regularization) to $\gamma = 1$ on the left (see main text for the definition of $\gamma$). Also 
plotted in both panels for comparison are $2\sigma$ power spectrum upper limits for the uniform-weighted case (blue) and inverse variance 
weighted case (black); both are after signal loss estimation. Finally, a theoretical prediction for noise ($2\sigma$ error) is plotted 
as green. In the PAPER-64 analysis in this paper, we choose to use a regularization scheme of $\widehat{\textbf{C}}_{\rm eff} \equiv 0.09 \, $Tr($\widehat{\textbf{C}})\textbf{I} + \widehat{\textbf{C}}$ ($\gamma = 0.09$) as a simple example of regularization that minimizes loss, and note that the power spectrum limits using this type of regularization are roughly constant across a large range of values of $\gamma$.}
	\label{fig:sigloss_modeloop}
\end{figure*}

In addition to our thermal noise prediction (green) and uniform-weighted power spectrum limit (blue), one additional horizontal line is shown in Figure \ref{fig:sigloss_modeloop} 
in both panels and represents a third regularization technique. This line (black) denotes the power spectrum value, post-signal loss estimation, for inverse variance weighting (multiplying an identity 
matrix element-wise to $\widehat{\textbf{C}}$). This result is single-valued and not a function of the horizontal axis. We see that all three regularization schemes shown (solid red top panel, solid red bottom panel, black) perform similarly at 
their best (i.e. when $\sim2$-$3$ eigenmodes are down-weighted in the case of the top panel's solid red curve). However, for the remainder of this paper, we choose to use the weighting option of $\widehat{\textbf{C}} \equiv \widehat{\textbf{C}} + 0.09 \,$Tr($\widehat{\textbf{C}})\textbf{I}$, or $\gamma = 0.09$, which we will denote as $\widehat{\textbf{C}}_{\rm eff}$. We choose this weighting scheme merely as a simple example of regularizing PAPER-64 covariances, noting that the power spectrum upper limit remains roughly constant for a broad range of values of $\gamma$. We also note that all of the regularizations described here perform similarly to the uniformly weighted case. However, we find that the regularization of empirical covariances does outperform the uniform weighted case for redshifts (and $k$-values) for which there are stronger foregrounds, as presented in Kolopanis et al. (\textit{submitted}).

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.4\textwidth]{plots/ps1_noise_add.pdf}
	\includegraphics[width=0.4\textwidth]{plots/ps2_noise_add.pdf}
	\caption{The power spectrum for a noise simulation that mimics the noise level of a subset of PAPER-64 data, where the solid red curve is the $2\sigma$ upper limit on the EoR signal estimated from our signal injection framework using $\widehat{\textbf{C}}_{\rm eff}$. The theoretical $2\sigma$ thermal noise level prediction based on observational parameters (calculated by Equation \eqref{eq:sense}) is in green. Additionally, the power spectrum results for the uniform weighted case is shown in three different ways: power spectrum values (black and gray points as positive and negative values, respectively, with $2\sigma$ error bars), the $2\sigma$ upper limit on the EoR signal using our full signal injection framework (solid blue), and the measured power spectrum values with $2\sigma$ thermal noise errors (gray shaded regions). The vertical dashed black lines signify the horizon limit for this analysis using $30$\,m baselines. We highlight that the bootstrapped data points and thermal noise prediction show good agreement, while the limits from the full injection framework (red and blue) are inflated due to the additional inclusion of sample variance that comes from the injection simulations.}
	\label{fig:ps_noise}
\end{figure*}

@@

Given our new understanding of the sensitivity of bootstraps to the number of elements sampled, we have removed the second 
bootstrapping step along time entirely and now simply bootstrap over the baseline axis. Power spectrum $2\sigma$ errors (computed from bootstrap variances) with this bootstrapping change for fringe-rate filtered noise are shown in Figure 
\ref{fig:data_errors}. The estimates are uniformly weighted in order to disentangle the effects of bootstrapping from signal loss. As 
shown in the figure, when more elements are drawn for each bootstrap than the number of 
independent samples (by over-sampling elements along the time axis), repeated values begin to crop up and the apparent variation between bootstraps drops, resulting in limits (gray) below the predicted noise level (green). Using the revised bootstrapping method, where bootstrapping only occurs over the baseline axis, the limits (black) are shown to better agree with the analytic prediction for noise. While Figure \ref{fig:data_errors} implies that errors are under-estimated by a factor of $\sim$ $5$ in mK$^{2}$ for the noise simulation, in practice this factor is lower for the case of real data (a factor of $\sim$ $3$ in mK$^{2}$ instead), possibly due to the data being less correlated in time than the fringe-rate filtered noise in the simulation. }%Figure \ref{fig:data_errors} implies that power spectrum upper limits were originally under-estimated by a factor of $\sim$ $5$ in mK$^{2}$.

\item{A second change to our bootstrapping procedure is that we now bootstrap over baseline cross-products, instead of the baselines themselves. In the previous analysis, baselines were bootstrapped prior to forming cross power spectra, and using this particular ordering of operations (bootstrapping, then cross-multiplication) yields variances that have been found to disagree with predicted errors from bootstrapping using simulations. On the contrary, bootstrapping over cross power spectra ensures that we are estimating the variance of our quantity of interest (i.e. the power spectrum). This change, while fundamental in retaining the integrity of the bootstrapping method in general, alters the resulting power spectrum errors by factors of $<2$ in practice.}

@@

The second null test possibility (error bars consistent with zero but not with thermal noise) is displayed by the $k$-values just outside the horizon ($k\sim\pm 0.1$\,$h$ Mpc$^{-1}$) for all three tests. This indicates an additive noise component that is increasing our errors. More specifically, although we expect each cross-multiplication that is used in power spectrum estimation to have independent noise, there is still the possibility of a noise-foreground coupling term that can introduce power. This is because cross-multiplications produce four additive terms --- a signal-squared term (where ``signal" includes both foregrounds and EoR), two cross-terms between the signal and noise, and one noise-only term. When differencing two power spectra (each with their own four terms), we expect the signal-only term to subtract out for a ``passing" null test, and we expect the noise-only terms to be consistent with thermal noise. While the cross-terms have a mathematical expectation value of zero, in practice we are limited by our number of samples ($90$ cross-products for this analysis times $\sim$ $8$ independent LST samples). Combined with the fact that the foregrounds are so bright, the finite ensemble of the couplings can introduce extra variance that varies with foreground strength. It is therefore not surprising that this effect is largest at $k$-values just outside the horizon, where we expect foregrounds to be brightest post-delay filtering.

@@

\begin{itemize}
\item Signal loss, previously found to be $<2\%$ in \citetalias{ali_et_al2015}, was under-estimated by a factor of $>$$1000$ for empirically estimated inverse 
covariance weighting. Using a regularized covariance weighting method can minimize loss (Section 
\ref{sec:Weight}), however, because a regularized weighting method is not as aggressive as the former, our results are still a 
factor of $\sim10$ in mK higher than the lossy empirical inverse covariance limits. Under-estimated signal loss therefore represents the bulk of our revision. 
This revision is similar in nature to the re-analysis of results from the GMRT (\citealt{paciga_et_al2013}) which were also 
revised from new signal loss calculations associated with their singular value decomposition foreground filter.
\item Power spectrum errors, originally computed by bootstrapping, were under-estimated for the data by a factor of $\sim2$ in mK due to oversampling data whose effective number of independent samples was reduced from fringe-rate filtering (Section \ref{sec:Boot}). Several other errors were also found regarding error estimation, though with smaller effects.
\item Several factors used in an analytic expression to predict the noise-level in PAPER-64 data were revised, yielding a 
decrease in predicted sensitivity level by a factor of $\sim3$ in mK (Section \ref{sec:PSSense}). We note that our sensitivity prediction is revised by a factor less than our overall
power spectrum result, implying that if taken at face value, the theoretical prediction for noise in \citetalias{ali_et_al2015} was too high for its data 
points.
\end{itemize}

@@

When we simulate our full injection framework as in Section \ref{sec:Practice}, we sample 50 $P_{\rm in}$ values that range from $\sim$ $10^{5}$\,mK$^{2}$ ($h^{-1}$ Mpc)$^{3}$ to $\sim$ 10$^{11}$\,mK$^{2}$ ($h^{-1}$ Mpc)$^{3}$, and we note that the prior is set to zero outside those regions. For the injections that we do sample, we can easily fit analytic functions to the mean and standard deviations of $\widehat{P}_{\rm out}$ ($\bar y$ and $\sigma$) as functions of $P_{\rm in}$. An example of the typical shape of these functions for the PAPER-64 analysis is shown in Figure \ref{fig:jeffreys1}, though in practice we fit solutions for every $k$-value and simulation independently.

@@

\begin{figure}
	\centering
	\includegraphics[width=10cm]{plots/jeffrey_prior.pdf}
	\caption{An example of the typical Jeffreys prior shape for the PAPER-64 analysis as computed by Equation \eqref{eq:jeffreys_final} (black). We smooth the prior using a sliding boxcar average over every $5$ injections (red). Most noticeably, the Jeffreys prior is not constant with $P_{\rm in}$, meaning a uniform prior would be an informative prior.}
	\label{fig:jeffreys2}
\end{figure}
